The Evolution of Phenotypically Invariant Genetic Networks

### 2017-02-03 kalman_walk.tex
started to add citations to the intro.

\documentclass[11 pt]{article}
\usepackage{amsmath, amssymb, color, xcolor}
\usepackage{graphicx}
\usepackage{fullpage}
\usepackage[hidelinks]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
\usepackage{authblk}
\newcommand{\plr}[1]{{\color{blue}\it #1}}
\newcommand{\jss}[1]{{\color{olive}\it #1}}
% \newcommand{\ddt}{\frac{d}{dt}}
\newcommand{\ddt}{\dot}
\newcommand{\ro}{{ro}}
\newcommand{\nro}{{\bar{r}o}}
\newcommand{\rno}{{r\bar{o}}}
\newcommand{\nrno}{{\bar{r}\bar{o}}}
\newcommand{\reachable}{\mathcal{R}}
\newcommand{\unobservable}{\bar{\mathcal{O}}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}

@@

Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery. Progress in a suite of biological subdisciplines is stalled by our general lack of understanding of this molecular machinery: with respect to both its function and evolution. There does exist a growing body of experiment and data on the evolutionary histories and molecular characterizations of particular gene regulatory networks, as well as thoughtful verbal and conceptual models. However, as Hardy and Weinberg taught us over a century ago, verbal theories are often insufficient, if not downright misleading. This is especially pertinent given the staggering complexity and scope of contemporary research programs. This outlook necessitates the advancement of conceptual frameworks of such precision, only mathematics will suffice. Here we outline an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identification, and dynamical systems theory. Presently, we focus on the neutral evolution of genetic regulatory networks. That is, we analytically describe the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them. In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe evolution under the ``conservation of phenotype'' as a Brownian motion over phenotypically-invariant network-space. This analysis provides insight to the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function.

It is commonly taught that an organism's genome contains the heritable material 
that natural selection filters and that an organism's phenotype directly determines its
evolutionary fitness. Between genotype and phenotype is an often complicated and poorly
undrstood molecular machinery -- and it is a major goal common to many different disciplines
within the life sciences to elucidate its form, function, and evolution. These aims are
delicately intertwined and a comphrehensive understanding of a system's evolution requires
an understanding of its function and vice versa.

The molecular machinery, interacting with the environment, and bridging genotype to phenotype
can be mathematically described as a dynamical system (or a system of differential equations).
Previously it has been argued that any idealized study of evolution is incomplete without
a mathematically sufficient description of the genotype, phenotype, and transformation from
one to the other. Movement in this direction is ongoing, as researhers have begun to study 
the evolution of empirically inspired computational and mathematical models of gene regulatory networks (GRNs) and metabolic networks that incorporate empirical data such as sequences and expression patterns. If we allow the reasonable assumption that the genotype-phenotype map can be represented as a system of differential equations, we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.

In some fields that seek to fit parametric models to experimental data, such as control
theory, chemical engineering, and statistics, it is well known that mathematical models
can fundamentally be \emph{unidentifiable} and/or \emph{indistinguishable} -- meaning that 
there can be uncertaintity about an inferred model's parameters or even its claims about
causal structure, even with access to complete and perfect data. Models with different 
parameter schemes, or even different mechanics can be equally accurate, but still not
\emph{actually} agree with what is being modelled. In control theory, where electrical 
circuits and mechanical systems are often the focus, it is understood that there can be an 
infinite number of ``realizations,'' or ways to reverse engineer the dynamics of a black box,
even if all possible input and output experiments on the black box are performed. In chemical
engineering, those who study chemical reaction networks sometimes refer to the fundamental
unidentifiability of these networks as ``the fundamental dogma of chemical kinetics.''
Although this may frusturate the occasional engineer or scientist, viewed from another angle,
the concepts of unidentifiability and indistinguishability can provide a starting point for
thinking about externally equivalent systems -- systems that evolution can explore, so long
as the parameters and structures can be realized biologically. In fact, evolutionary
biologists who study homology and analogy are very familiar with such functional symmetries; macroscopically identical phenotypes in closely related species can be divergent at the molecular level.

@@

What are typically referred to as ``design principles'' or some sort of clear structure-function relationships are commonly saught

Seeking to attribute functionality, uninformed by evolution, can lead to spurious claims
(i.e. see Graur).

@@

Organisms often have to respond to their environments,
and the way this happens is that external input triggers a cascade of molecular signals
whose eventual result is a reaction,
whose approprateness for the situation determines fitness.
\plr{A similar situation occurs in development (...?).}
However, there are many ways to skin a cat:
especially given a number of possible molecular intermediaries,
there may be a large number of ways to construct such a molecular signalling system
that has precisely the \emph{same} input-output relationships,
and hence the same function.
Therefore, any mutation that changes one such system to another, equivalent version
will be neutral,
and may well become polymorphic in a population if there are no ill effects in heterozygotes.
Certainly, mutations that move a system away from optimal functioning will also appear,
but in very large populations will not rise to high frequency.
\plr{Deleterious things are quite important but somehow justify starting with the neutral ones.}

>>>>

\documentclass[11 pt]{article}
\usepackage{amsmath, amssymb, color, xcolor}
\usepackage{graphicx}
\usepackage{fullpage}
\usepackage[backref, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
\usepackage{authblk}
\newcommand{\plr}[1]{{\color{blue}\it #1}}
\newcommand{\jss}[1]{{\color{olive}\it #1}}
% \newcommand{\ddt}{\frac{d}{dt}}
\newcommand{\ddt}{\dot}
\newcommand{\ro}{{ro}}
\newcommand{\nro}{{\bar{r}o}}
\newcommand{\rno}{{r\bar{o}}}
\newcommand{\nrno}{{\bar{r}\bar{o}}}
\newcommand{\reachable}{\mathcal{R}}
\newcommand{\unobservable}{\bar{\mathcal{O}}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}

@@

Bridging the gulf between an organism's genome and phenotype is a poorly understood and complex molecular machinery. Progress in a suite of biological subdisciplines is stalled by our general lack of understanding of this molecular machinery: with respect to both its function and evolution. There does exist a growing body of experiment and data on the evolutionary histories and molecular characterizations of particular gene regulatory networks \citep{jaeger2011gap, davidson2006gene, israel2016comparative}, as well as thoughtful verbal and conceptual models \citep{true2001developmental, gwagner1}. However, as Hardy and Weinberg taught us over a century ago, verbal theories are often insufficient, if not downright misleading \citep{hardy1908mendelian, weinberg1908vererbungsgesetze}. This is especially pertinent given the staggering complexity and scope of contemporary research programs. This outlook necessitates the advancement of conceptual frameworks of such precision, only mathematics will suffice. Previously it has been suggested that any idealized study of evolution is incomplete without a mathematically sufficient description of the genotype, phenotype, and transformation from one to the other \citep{Lewontin1974genetic}.%Here we outline an analytical theory to study the evolution of biological systems, borrowing insight and methods from control engineering, systems identification, and dynamical systems theory.

%Presently, we focus on the neutral evolution of genetic regulatory networks. That is, we analytically describe the set of all linear gene networks (of any size) that produce identical phenotypes -- and the evolutionary paths connecting them. In the idealized case of a perfectly adapted population, constant selection, and a static environment, we observe evolution under the ``conservation of phenotype'' as a Brownian motion over phenotypically-invariant network-space. This analysis provides insight to the mechanisms and parameters important for understanding developmental systems drift, network rewiring, evolvability, epistasis, and speciation, as well as the tenuous connection between network architecture and function.

%It is commonly taught that an organism's genome contains the heritable material 
%that natural selection filters and that an organism's phenotype directly determines its
%evolutionary fitness. Between genotype and phenotype is an often complicated and poorly
%undrstood molecular machinery -- and it is a major goal common to many different disciplines
%within the life sciences to elucidate its form, function, and evolution. These aims are
%delicately intertwined and a comphrehensive understanding of a system's evolution requires
%an understanding of its function and vice versa.

The molecular machinery, interacting with the environment, and bridging genotype to phenotype
can be mathematically described as a dynamical system -- or a system of differential equations\citep{jaeger2015comet}.
 Movement in this direction is ongoing, as researchers have begun to study 
the evolution of both abstract \citep{wagner1994evolution, wagner1996does,  siegal2002waddington, bergman2003evolutionary} and empirically inspired computational and mathematical models of gene regulatory networks (GRNs) \citep{mjolsness1991connectionist, jaeger2004dynamic, maria1, vitaly1, vitaly2, crombach2016gap, wotton2015quantitative}. If we allow the reasonable assumption that the genotype-phenotype map can be represented as a system of differential equations, we can immediately discuss its evolution and function in a much more mechanistic, yet general, manner.

In some fields that seek to fit parametric models to experimental data, such as control
theory, chemical engineering, and statistics, it is well known that mathematical models
can fundamentally be \emph{unidentifiable} and/or \emph{indistinguishable} -- meaning that 
there can be uncertaintity about an inferred model's parameters or even its claims about
causal structure, even with access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. Models with different 
parameter schemes, or even different mechanics can be equally accurate, but still not
\emph{actually} agree with what is being modelled. In control theory, where electrical 
circuits and mechanical systems are often the focus, it is understood that there can be an 
infinite number of ``realizations,'' or ways to reverse engineer the dynamics of a black box,
even if all possible input and output experiments on the black box are performed \citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. In chemical
engineering, those who study chemical reaction networks sometimes refer to the fundamental
unidentifiability of these networks as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.
Although this may frusturate the occasional engineer or scientist, viewed from another angle,
the concepts of unidentifiability and indistinguishability can provide a starting point for
thinking about externally equivalent systems -- systems that evolution can explore, so long
as the parameters and structures can be realized biologically. In fact, evolutionary
biologists who study homology and analogy are very familiar with such functional symmetries; macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level \citep{true2001developmental, tsong2006evolution, hare2008sepsid, vierstra2014mouse, stergachis2014conservation, taylor2016diverse, matsui2015regulatory}.

@@

What are typically referred to as ``design principles'' or some sort of clear structure-function relationships are commonly saught \citep{milo2002network, alon2007network, ma2009defining}.

Seeking to attribute functionality, uninformed by evolution, can lead to spurious claims
(i.e. see Graur \citep{lynch2007evolution, lynch2007frailty, graur2013immortality}).

@@

Organisms often have to respond to their environments,
and the way this happens is that external input triggers a cascade of molecular signals
whose eventual result is a reaction,
whose approprateness for the situation determines fitness.
\plr{A similar situation occurs in development (...?).}
However, there are many ways to skin a cat \jss{(there are also many ways to develop a cat)}:
especially given a number of possible molecular intermediaries,
there may be a large number of ways to construct such a molecular signalling system
that has precisely the \emph{same} input-output relationships,
and hence the same function.
Therefore, any mutation that changes one such system to another, equivalent version
will be neutral,
and may well become polymorphic in a population if there are no ill effects in heterozygotes.
Certainly, mutations that move a system away from optimal functioning will also appear,
but in very large populations will not rise to high frequency.
\plr{Deleterious things are quite important but somehow justify starting with the neutral ones.}
